Parametricity for Primitive Nested Types and GADTs

### 2020-12-16 free-theorems-nesteds/20-mscs.tex
Sections 5-8 for proofreading

Note that since the body F of a type $(\mu \phi. \lambda
\ol\alpha. F)\ol{G}$ can only be functorial in $\phi$ and the
variables in $\ol{\alpha}$, the representation of
$\mathit{List}\,\alpha$ as the ADT $\mu \beta. \,\onet + \alpha \times
\beta$ cannot be functorial in $\alpha$. By contrast, if
$\mathit{List}\,\alpha$ is represented as the nested type $(\mu
\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ then
we can choose $\alpha$ to be a functorial variable or not when forming
the type. This observation holds for other ADTs as well; for example,
if $\mathit{Tree}\,\alpha\,\gamma = \mu \beta. \alpha + \beta \times
\gamma \times \beta$, then $\alpha, \gamma; \emptyset \vdash
\mathit{Tree}\,\alpha\,\gamma$ is well-formed, but $\emptyset; \alpha,
\gamma \vdash \mathit{Tree}\,\alpha\,\gamma$ is not. And it also
applies to some non-ADT types, such as $\mathit{GRose}\,\phi\,\alpha =
\mu \beta. \onet + \alpha \times \phi\beta$, in which $\phi$ and
$\alpha$ must both be non-functorial variables. {\color{red} It is in
  fact possible to allow ``extra'' $0$-ary functorial variables in the
  body of $\mu$-types (functorial variables of higher arity are the
  real problem), which would allow the first-order representations of
  ADTs to be functorial.  However, it simplifies the formation rule
  for $\mu$-types considerably to disallow even these $0$-ary
  functorial variables. Since we can always use an ADT's (semantically
  equivalent) second-order representation when functoriality is
  needed, this restriction allows us to simplify the type syntax
  without sacrificing the expressivity of the calculus.}
%The representation of $\mathit{List}\,\alpha$ as the ADT $\mu
%\beta. \onet + \alpha \times \beta$ cannot be functorial in
%$\alpha$. By contrast, its representation as the nested type $(\mu
%\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ can be
%considered either functorial in $\alpha$ or not, according to whether
%$\alpha$ is in the functorial or non-functorial context in which this
%type is formed. If $\emptyset;\alpha \vdash \mathit{List} \,\alpha$,
%then $\vdash \Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is
%well-formed; if $\alpha;\emptyset \vdash \List \,\alpha$, then it is
%not.  Similarly, if $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\,
%\alpha + \beta \times \gamma \times \beta$, then $\gamma;\emptyset
%\vdash \Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \,
%\alpha\, \gamma)$ is well-formed, but $\emptyset;\gamma \vdash
%\Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
%\gamma)$ is not. The body $F$ of a type $(\mu \phi.\lambda
%\ol\alpha.F)\ol G$ can only be functorial in $\phi$ and the variables
%in $\ol\alpha$. As a result, both $\phi$ and the variables in
%$\ol\alpha$ must be non-functorial in $\mathit{GRose}\, \phi \, \alpha
%= \mu\beta . \alpha \times \phi \beta$, $\mathit{GRose}\,\phi\,\alpha$
%cannot be the (co)domain of a $\Nat$-type binding any variables, and
%no type of the form $(\mu \phi. \lambda
%\alpha. \mathit{GRose}\,\phi\,\alpha)\tau$ is well-formed in our
%system.
These restrictions ensure that the set and relational interpretations
of types in Section~\ref{sec:type-interp} are functors. In
Section~\ref{sec:conclusion} we discuss how more expressive types
might be accommodated.

@@

We first show that terms of $\Nat$-type behave as natural
transformations with respect to their source and target functorial
types. If $\Gamma; \ol{\alpha} \vdash F$ then define the {\em
  identity} $\id_F$ on $F$ by $\id_F = \Gamma;\emptyset~|~\emptyset
\vdash L_{\ol{\alpha}}x.x : \Nat^{\ol{\alpha}} F\,F$, and if $\Gamma;
\emptyset \,|\, \Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and
$\Gamma; \emptyset \,|\, \Delta \vdash s: \Nat^{\overline{\alpha}}
G\,H$ are terms then define the {\em composition} $s \circ t$ of $t$
and $s$ by $s \circ t = \Gamma; \emptyset\,|\, \Delta \vdash
L_{\overline{\alpha}}
x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
\Nat^{\overline{\alpha}} F\,H$. Then \[\setsem{\Gamma; \emptyset \,|\,
  \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
\id_{\lambda \ol{A}. \setsem{\Gamma; \ol{\alpha} \vdash F} \rho
  [\ol{\alpha := A}]}\] for any set environment $\rho$, and
\[\setsem{\Gamma; \emptyset \,|\, \Delta \vdash s \circ t:
  \Nat^{\overline{\alpha}} F\,H} = \setsem{\Gamma; \emptyset \,|\,
  \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H} \circ
\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
  \Nat^{\overline{\alpha}} F\,G}\]
Naturality of term interpretations is then easily verified:
\begin{thm}\label{eq:ft-from-nat} 
  If\, $\Gamma; \emptyset \,|\, \Delta \vdash s : \Nat^{\overline{\alpha},
  \overline{\gamma}} F\,G$ and $\overline{\Gamma; \emptyset \,|\,
  \Delta \vdash t : \Nat^{\overline{\gamma}} K\, H}$, then
  \[\begin{array}{l}
  \hspace*{0.135in}
\setsem{\Gamma; \emptyset\,|\, \Delta
  \vdash 
  ((\map_G^{\overline{K}, \overline{H}})_\emptyset \,\overline{t}) \circ
(L_{\overline{\gamma}} z. s_{\overline{K}, \overline{\gamma}}
  z)
  : \Nat^{\overline{\gamma}} F[\overline{\alpha := K}]\,
  G[\overline{\alpha := H}]}\\
= \setsem{ \Gamma; \emptyset \,|\, \Delta \vdash
(L_{\overline{\gamma}} z.
  s_{\overline{H}, \overline{\gamma}} z)\circ
  ((\map_F^{\overline{K}, \overline{H}})_\emptyset \,
  \overline{t})  : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
      K}]\, G[\overline{\alpha := H}]}
\end{array}\]
\end{thm}
\noindent

>>>>

Note that since the body F of a type $(\mu \phi. \lambda
\ol\alpha. F)\ol{G}$ can only be functorial in $\phi$ and the
variables in $\ol{\alpha}$, the representation of
$\mathit{List}\,\alpha$ as the ADT $\mu \beta. \,\onet + \alpha \times
\beta$ cannot be functorial in $\alpha$. By contrast, if
$\mathit{List}\,\alpha$ is represented as the nested type $(\mu
\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ then
we can choose $\alpha$ to be a functorial variable or not when forming
the type. This observation holds for other ADTs as well; for example,
if $\mathit{Tree}\,\alpha\,\gamma = \mu \beta. \alpha + \beta \times
\gamma \times \beta$, then $\alpha, \gamma; \emptyset \vdash
\mathit{Tree}\,\alpha\,\gamma$ is well-formed, but $\emptyset; \alpha,
\gamma \vdash \mathit{Tree}\,\alpha\,\gamma$ is not. And it also
applies to some non-ADT types, such as $\mathit{GRose}\,\phi\,\alpha =
\mu \beta. \onet + \alpha \times \phi\beta$, in which $\phi$ and
$\alpha$ must both be non-functorial variables.  It is in fact
possible to allow ``extra'' $0$-ary functorial variables in the body
of $\mu$-types (functorial variables of higher arity are the real
problem), which would allow the first-order representations of ADTs to
be functorial. However, doing this requires some changes to the
formation rule for $\mu$-types, as well as the delicate threading of
some additional conditions throughout the model construction.  But
since we can always use an ADT's (semantically equivalent)
second-order representation when functoriality is needed, disallowing
such ``extra'' variables does not negatively impact the expressivity
of our calculus. We therefore pursue the simpler syntax here.
%The representation of $\mathit{List}\,\alpha$ as the ADT $\mu
%\beta. \onet + \alpha \times \beta$ cannot be functorial in
%$\alpha$. By contrast, its representation as the nested type $(\mu
%\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ can be
%considered either functorial in $\alpha$ or not, according to whether
%$\alpha$ is in the functorial or non-functorial context in which this
%type is formed. If $\emptyset;\alpha \vdash \mathit{List} \,\alpha$,
%then $\vdash \Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is
%well-formed; if $\alpha;\emptyset \vdash \List \,\alpha$, then it is
%not.  Similarly, if $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\,
%\alpha + \beta \times \gamma \times \beta$, then $\gamma;\emptyset
%\vdash \Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \,
%\alpha\, \gamma)$ is well-formed, but $\emptyset;\gamma \vdash
%\Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
%\gamma)$ is not. The body $F$ of a type $(\mu \phi.\lambda
%\ol\alpha.F)\ol G$ can only be functorial in $\phi$ and the variables
%in $\ol\alpha$. As a result, both $\phi$ and the variables in
%$\ol\alpha$ must be non-functorial in $\mathit{GRose}\, \phi \, \alpha
%= \mu\beta . \alpha \times \phi \beta$, $\mathit{GRose}\,\phi\,\alpha$
%cannot be the (co)domain of a $\Nat$-type binding any variables, and
%no type of the form $(\mu \phi. \lambda
%\alpha. \mathit{GRose}\,\phi\,\alpha)\tau$ is well-formed in our
%system.
%These restrictions ensure that the set and relational interpretations
%of types in Section~\ref{sec:type-interp} are functors. In
%Section~\ref{sec:conclusion} we discuss how more expressive types
%might be accommodated.

@@

We first show that terms of $\Nat$-type behave as natural
transformations with respect to their source and target functorial
types, and derive some conseqences of this observation. If $\Gamma;
\ol{\alpha} \vdash F$ then define the {\em identity} $\id_F$ on $F$ by
$\id_F = \Gamma;\emptyset~|~\emptyset \vdash L_{\ol{\alpha}}x.x :
\Nat^{\ol{\alpha}} F\,F$, and if $\Gamma; \emptyset \,|\, \Delta
\vdash t: \Nat^{\overline{\alpha}} F\,G$ and $\Gamma; \emptyset \,|\,
\Delta \vdash s: \Nat^{\overline{\alpha}} G\,H$ are terms then define
the {\em composition} $s \circ t$ of $t$ and $s$ by $s \circ t =
\Gamma; \emptyset\,|\, \Delta \vdash L_{\overline{\alpha}}
x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
\Nat^{\overline{\alpha}} F\,H$. Then \[\setsem{\Gamma; \emptyset \,|\,
  \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
\id_{\lambda \ol{A}. \setsem{\Gamma; \ol{\alpha} \vdash F} \rho
  [\ol{\alpha := A}]}\] for any set environment $\rho$, and
\[\setsem{\Gamma; \emptyset \,|\, \Delta \vdash s \circ t:
  \Nat^{\overline{\alpha}} F\,H} = \setsem{\Gamma; \emptyset \,|\,
  \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H} \circ
\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
  \Nat^{\overline{\alpha}} F\,G}\]
Naturality of term interpretations is then easily verified:
\begin{thm}\label{eq:ft-from-nat} 
  If\, $\Gamma; \emptyset \,|\, \Delta \vdash s : \Nat^{\overline{\alpha},
  \overline{\gamma}} F\,G$ and $\overline{\Gamma; \emptyset \,|\,
  \Delta \vdash t : \Nat^{\overline{\gamma}} K\, H}$, then
  \[\begin{array}{l}
  \hspace*{0.135in}
\setsem{\Gamma; \emptyset\,|\, \Delta
  \vdash 
  ((\map_G^{\overline{K}, \overline{H}})_\emptyset \,\overline{t}) \circ
(L_{\overline{\gamma}} z. s_{\overline{K}, \overline{\gamma}}
  z)
  : \Nat^{\overline{\gamma}} F[\overline{\alpha := K}]\,
  G[\overline{\alpha := H}]}\\
= \setsem{ \Gamma; \emptyset \,|\, \Delta \vdash
(L_{\overline{\gamma}} z.
  s_{\overline{H}, \overline{\gamma}} z)\circ
  ((\map_F^{\overline{K}, \overline{H}})_\emptyset \,
  \overline{t})  : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
      K}]\, G[\overline{\alpha := H}]}
\end{array}\]
\end{thm}
\noindent
